Cognate beginnings to bilingual lexical acquisition

### 2023-08-12 manuscript/manuscript.qmd
Fix coefficients table

One important observation of studies on bilinguals' early vocabulary acquisition is that cognate words are easier to acquire than non-cognate words. Cognate words are translation equivalents that are phonologically similar (or share some type of form-similarity). For instance, the Spanish translation equivalent of *cat* is *gato*, a cognate word; the translation equivalent of *dog* is *perro*, a non-cognate word. The differences in the percentages of cognate and non-cognate words between two languages is related to historical reasons: languages typologically close (like Dutch and English or Italian and Spanish) share more cognates than languages typologically distant (like English and Chinese, or Urdu and Spanish). The conclusion that cognate words are easier to learn is based on two types of evidence: studies investigating vocabulary sizes in children learning language pairs with different percentages of cognates (that is, differing in their typological distance) and studies comparing the number of cognate and non-cognate words children know in a specific language pair.

@@

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts. Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @hoshino2008cognate; @morford2011when; @spivey1999cross; @shook2012bimodal; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @floccia2020translation; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @poulin-dubois2013lexical].

Although there is a consensus on the role of parallel activation in bilinguals' lexical processing and acquisition, previous studies do not address its influence on the learning trajectories of words. Results are aggregated across words and provide no information about the specific dynamics of how parallel activation influences word learning. This is the goal of the present research.

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated. We use the term *learning instance* in the fashion of accumulator models of language acquisition; as an exposure to a word-form that constitutes an opportunity for the child to accumulate information about the word. We do not assume if a learning instance is a discrete or a continuous unit of accumulation of information. We consider that a learning instance of a word consists of an exposure to the its phonological form if the resulting strength of activation of its phonological representation in the child's lexicon reaches some thoretical threshold that leads to word-form recognition. This activation may result from the infant being exposed to the actual word-form, or the result of activation spreading through phonological or semantic links across lexical representation, as in the case of parallel activation. The strength of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more strongly than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language would receive stronger activation from words in the higher-exposure language than vice versa. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language. This asymmetric cross-language activation would be consistent with previous reports of larger priming effects from the dominant to the non-dominant language.

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire (BVQ) [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formr platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory (OCDI) [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic or functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for *dog*), roughly following a one-to-one mapping. Some of the words in Catalan did not have a clear translation or had more than one possible translation in Spanish, and vice versa, therefore the unequal number of words included in the two lists (see @tbl-items for a summary of the questionnaire items).

@@

For each word included in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Given the large number of words in the vocabulary checklists, we created four different subsets of the complete list of items. Each subset contained a random but representative sub-sample of the items from the complete list (see @tbl-items). Semantic or functional categories with less than 16 items---thus resulting in less than four items after dividing it in four lists---were not divided in the short version of the questionnaire: all of their items were included in the four lists. Items that were part of the trial lists of some ongoing experiments in the lab were also included in all versions. The resulting reduced list contained between `r min(n_items_version$catalan)` and  `r max(n_items_version$catalan)` Catalan words, and between `r min(n_items_version$spanish)` and  `r max(n_items_version$spanish)` Spanish words. Participants were randomly allocated into one of the four subsets.

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between the `r format(dates[1], "%dth of %B, %Y")` and the `r format(dates[2], "%dth of %B, %Y")`: `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)`, and were always allocated to the same questionnaire list (A, B, C, or D). Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra) and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r scales::percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

@@

We collected data for `r format(nrow(n_items), big.mark = ",")` words. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`, see @sec-predictors), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together. Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the *tidyverse* family of packages [@wickham2019welcome].

@@

We developed the *Exposure* predictor to account for the fact that bilinguals' exposure to a given word-form is not only a function of the word-form's lexical frequency, but also of the quantitative input they receive from the language such word-form belongs to, we expressed lexical frequencies as the product between both variables. First, we extracted the child-directed lexical frequency of each word-form from the CHILDES database [@macwhinney2000childes; @sanchez2019childesdb]. Using the corresponding lexical frequencies directly from Catalan and Spanish was not possible due to the low number of Catalan participants and tokens available in their corresponding CHILDES corpora, so they were extracted from the English corpora instead. We mapped the lexical frequencies of the English words to their Catalan and Spanish translations [see @fourtassi2020growth for a similar approach], and transformed them to Zipf scores [@vanheuven2014subtlexuk; @zipf1949human]. We multiplied the resulting lexical frequencies by the reported degree of exposure of the child to Catalan or Spanish. For instance, for a child whose degree of exposure is 80% for Catalan and 20% for Spanish, the expected *Exposure* score to the Catalan word-form *cotxe* [*car*]---with a lexical frequency of `r round(items$freq[items$item=="cotxe"], 2)`---would be `r round(items$freq[items$item=="cotxe"]*0.80, 2)`, while that of its translation to Spanish *coche* would be `r round(items$freq[items$item=="coche"]*0.20, 2)`.

We defined *Cognateness* as the phonological similarity between each word-form and its translation. For each translation equivalent, we used the `stringdist` [@vanderloo2014stringdist] R package to calculate the Levenshtein distance between the Catalan and the Spanish phonological transcriptions of the word-forms. The Levenshtein distance measures the number of editions (insertions, deletions, or substitutions) that one string of characters must go through to become identical to the other [@levenshtein1966binary]. We divided the Levenshtein distance of each translation equivalent by the length of the longest word-form to correct for word-form length (longer strings are likely to show a larger number of mismatches). Finally, we subtracted the result from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different (e.g., /\textipa{"taw.l5}/--/\textipa{"me.sa}/, *table*), and one indicates that the two word-forms are identical (e.g., /\textipa{"mar}/--/\textipa{"mar}/, *sea*) [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; for similar approaches]. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor.

@@

post_draws_list <- model_summary |>
    mutate(across(
        .value:.lower,
        \(x) ifelse(grepl("Intercept", .type),
            plogis(x), x / 4
        )
    )) |>
    split(model_summary$.variable)

@@

The coefficient of *Age* showed the strongest association with the probability of acquisition ($\beta$ = `r round(post_draws_list[["b_age_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_age_std"]]$.lower, 3)`, `r round(post_draws_list[["b_age_std"]]$.upper, 3)`]), with all posterior samples falling out of the ROPE. A one-month increment in age increased a maximum of `r round(post_draws_list[["b_age_std"]]$.value/sd(responses$age), 2)` the probability of acquisition. Similarly, the word-form exposure index (*Exposure*) had a strong effect on the probability of acquisition ($\beta$ = `r round(post_draws_list[["b_exposure_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_exposure_std"]]$.lower, 3)`, `r round(post_draws_list[["b_exposure_std"]]$.upper, 3)`]). All of the posterior samples of this regression coefficient excluded the ROPE. The impact of this predictor on the probability of acquisition was positive: for every standard deviation increase in exposure, the participant was `r round(post_draws_list[["b_exposure_std"]]$.value/sd(responses$exposure), 3)` more likely to acquire it. Word-form length also showed a significant association with probability of acquisition ($\beta$ = `r round(post_draws_list[["b_n_phon_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_n_phon_std"]]$.lower, 3)`, `r round(post_draws_list[["b_n_phon_std"]]$.upper, 3)`]). For every phoneme in the word-form, participants were `r round((post_draws_list[["b_n_phon_std"]]$.value)/sd(responses$n_phon), 3)` less likely to know it. The 95% HDI of the regression coefficient of the *Age* $\times$ *Exposure* interaction also excluded the ROPE ($\beta$ = `r round(post_draws_list[["b_age_std:exposure_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_age_std:exposure_std"]]$.lower, 3)`, `r round(post_draws_list[["b_age_std:exposure_std"]]$.upper, 3)`]), showing that the effect of the word-form exposure index differed across ages: older children were more likely to acquire words with a higher exposure rate than younger children.

The posterior distribution of the main effect of cognateness excluded the ROPE completely ($\beta$ = `r round(post_draws_list[["b_lv_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_lv_std"]]$.lower, 3)`, `r round(post_draws_list[["b_lv_std"]]$.upper, 3)`]). For every 10% increment in cognateness, the acquisition of a word increased in `r round(post_draws_list[["b_lv_std"]]$.value/(40*sd(responses$lv)), 3)`. The effect of *Cognateness* interacted with that of *Exposure*: the 95% HDI of the regression coefficient of interaction excluded the ROPE entirely ($\beta$ = `r round(post_draws_list[["b_exposure_std:lv_std"]]$.value, 2)`, 95% HDI = [`r round(post_draws_list[["b_exposure_std:lv_std"]]$.lower, 3)`, `r round(post_draws_list[["b_exposure_std:lv_std"]]$.upper, 3)`]), suggesting that the effect of cognateness on a word's probability of acquisition changed depending on participants' exposure to the word-form. Follow-up analyses on this interaction showed that when exposure rate was low (e.g., -1 SD), cognateness increased the probability of acquisition substantially. This effect was negligible for words with median or high exposure (+1 SD) (see @fig-marginal).

@@

This study investigated the impact of cognateness (i.e., phonological similarity between translation equivalents) on the early bilingual lexicon. We used Bayesian Item Response Theory to model the acquisition trajectories of a large sample of Catalan and Spanish words, estimating the effect of cognateness on the probability of acquisition. This model corrected for participants' age, word-form length (number of phonemes), and a novel measure of participants' exposure rate to each word-form. Exposure rates were calculated as a language exposure-weighted lexical frequency score in which each word-form's lexical frequency was corrected by the degree to which the participant was exposed to each language. Overall, we found that cognates (i.e., phonologically similar translation equivalents) were acquired earlier than non-cognates. This effect was mediated by exposure rate: low-exposure word-forms benefited from their cognate status, whereas high-exposure word-forms did not. Using the concept of accumulator [see @kachergis2022standard for review], we provide a theoretical account of bilingual lexical acquisition. In the present account, parallel activation of the two languages plays a central role during the acquisition of early representations in the bilingual lexicon, and in which the dynamics of co-activation between translation equivalents results in an earlier age-of-acquisition.

>>>>

One important observation of studies on bilinguals' early vocabulary acquisition is that cognate words are easier to acquire than non-cognate words. Cognate words are translation equivalents that are phonologically similar (or share some type of form-similarity). For instance, the Spanish translation equivalent of *cat* is *gato*, a cognate word; the translation equivalent of *dog* is *perro*, a non-cognate word. For historical reasons, some pairs of languages share more cognates than others: languages typologically close (like Dutch and English or Italian and Spanish) share more cognates than languages typologically distant (like English and Chinese, or Urdu and Spanish). The conclusion that cognate words are easier to learn is based on two types of evidence: studies investigating vocabulary sizes in children learning language pairs with different percentages of cognates (that is, differing in their typological distance) and studies comparing the number of cognate and non-cognate words children know in a specific language pair.

@@

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts. Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @hoshino2008cognate; @morford2011when; @spivey1999cross; @shook2012bimodal; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*). Obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. More recntly, evidence of parallel activation has been reported in bilingual toddlers and children too [@vonholzen2012language; @floccia2020translation; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @poulin-dubois2013lexical]. Although there is a consensus on the role of parallel activation in bilinguals' lexical processing and acquisition, previous studies do not address its influence on the learning trajectories of words. Results are aggregated across words and provide no information about the specific dynamics of how parallel activation influences word learning. This is the goal of the present research.

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated. We use the term *learning instance* in the fashion of accumulator models of language acquisition; as an exposure to a word-form that constitutes an opportunity for the child to accumulate information about the word. We do not assume if a learning instance is a discrete or a continuous unit of accumulation of information. We consider that a learning instance of a word is an exposure to its (phonological) form if the resulting strength of activation of its representation in the child's lexicon reaches some thoretical threshold that leads to word-form recognition. This activation may result from the infant being exposed to the actual word-form, or the result of activation spreading through phonological or semantic links across lexical representation, as in the case of parallel activation. The strength of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more strongly than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language would receive stronger activation from words in the higher-exposure language than vice versa. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language. This asymmetric cross-language activation would be consistent with previous reports of larger priming effects from the dominant to the non-dominant language.

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire (BVQ) [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the `formr` platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic or functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for *dog*), roughly following a one-to-one mapping. Some of the words in Catalan did not have a clear translation or had more than one possible translation in Spanish, and vice versa, therefore the unequal number of words included in the two lists (see @tbl-items for a summary of the questionnaire items).

@@

For each word included in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Given the large number of words in the vocabulary checklists, we created four different subsets of the complete list of items: A, B, C, and D. Each subset contained a random but representative sub-sample of the items from the complete list (see @tbl-items). Semantic or functional categories with less than 16 items---thus resulting in less than four items after dividing it in four lists---were not divided: all of their items were included in the four lists. Items that were part of the trial lists of some ongoing experiments in the lab were also included in all versions. The resulting reduced list contained between `r min(n_items_version$catalan)` and  `r max(n_items_version$catalan)` Catalan words, and between `r min(n_items_version$spanish)` and  `r max(n_items_version$spanish)` Spanish words. Participants were randomly allocated into one of the four subsets.

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between Marth 30th, 2020 and October 31th, 2022: `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)` days, and were always allocated to the same questionnaire list (A, B, C, or D). Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra) and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r scales::percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed in less than 10% to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

@@

We collected data for `r format(nrow(n_items), big.mark = ",")` words. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`, see @sec-predictors), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together. Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`].

@@

We developed the *Exposure* predictor to account for the fact that bilinguals' exposure to a given word-form is not only a function of the word-form's lexical frequency, but also of the quantitative input they receive from the language such word-form belongs to. We expressed lexical frequencies as the product between both variables. First, we extracted the child-directed lexical frequency of each word-form from the CHILDES database [@macwhinney2000childes; @sanchez2019childesdb]. Using the corresponding lexical frequencies directly from Catalan and Spanish was not possible due to the low number of Catalan participants and tokens available in their corresponding CHILDES corpora, so they were extracted from the English corpora instead. We mapped the lexical frequencies of the English words to their Catalan and Spanish translations [see @fourtassi2020growth for a similar approach], and transformed them to Zipf scores [@vanheuven2014subtlexuk; @zipf1949human]. We multiplied the resulting lexical frequencies by the reported degree of exposure of the child to Catalan or Spanish. For instance, for a child whose degree of exposure is 80% for Catalan and 20% for Spanish, the expected *Exposure* score to the Catalan word-form *cotxe* [*car*]---with a lexical frequency of `r round(items$freq[items$item=="cotxe"], 2)`---would be `r round(items$freq[items$item=="cotxe"]*0.80, 2)`, while that of its translation to Spanish *coche* would be `r round(items$freq[items$item=="coche"]*0.20, 2)`.

We defined *Cognateness* as the phonological similarity between each word-form and its translation. For each translation equivalent, we used the `stringdist` [@vanderloo2014stringdist] R package to calculate the Levenshtein distance between the Catalan and the Spanish phonological transcriptions of the word-forms. The Levenshtein distance measures the number of editions (insertions, deletions, or substitutions) that one string of characters must go through to become identical to the other [@levenshtein1966binary]. We divided the Levenshtein distance of each translation equivalent by the length of the longest word-form to correct for word-form length (longer strings are likely to show a larger number of mismatches). Finally, we subtracted the result from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different (e.g., /\textipa{"taw.l5}/--/\textipa{"me.sa}/, *table*), and one indicates that the two word-forms are identical (e.g., /\textipa{"mar}/--/\textipa{"mar}/, *sea*) [@floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth]. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor.

@@

post_draws_list <- model_summary |>
    mutate(across(
        .value:.upper,
        \(x) ifelse(grepl("Intercept", .type),
            plogis(x), x / 4
        )
    )) |>
    split(model_summary$.variable)

@@

The coefficient of *Age* showed the strongest association with the probability of acquisition ($\beta$ = `r round(post_draws_list[["b_age_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_age_std"]]$.lower, 3)`, `r round(post_draws_list[["b_age_std"]]$.upper, 3)`]), with all posterior samples falling out of the ROPE. A one-month increment in age increased a maximum of `r round(post_draws_list[["b_age_std"]]$.value/sd(responses$age), 2)` the probability of acquisition. Similarly, the word-form exposure index (*Exposure*) had a strong effect on the probability of acquisition ($\beta$ = `r round(post_draws_list[["b_exposure_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_exposure_std"]]$.lower, 3)`, `r round(post_draws_list[["b_exposure_std"]]$.upper, 3)`]). All of the posterior samples of this regression coefficient excluded the ROPE. The impact of this predictor on the probability of acquisition was positive: for every standard deviation increase in exposure, the participant was `r round(post_draws_list[["b_exposure_std"]]$.value/sd(responses$exposure), 3)` probability points more likely to acquire it. Word-form length also showed a significant association with probability of acquisition ($\beta$ = `r round(post_draws_list[["b_n_phon_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_n_phon_std"]]$.lower, 3)`, `r round(post_draws_list[["b_n_phon_std"]]$.upper, 3)`]). For every phoneme in the word-form, participants were `r round((post_draws_list[["b_n_phon_std"]]$.value)/sd(responses$n_phon), 3)` probability points less likely to know it. The 95% HDI of the regression coefficient of the *Age* $\times$ *Exposure* interaction also excluded the ROPE ($\beta$ = `r round(post_draws_list[["b_age_std:exposure_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_age_std:exposure_std"]]$.lower, 3)`, `r round(post_draws_list[["b_age_std:exposure_std"]]$.upper, 3)`]), showing that the effect of the word-form exposure index differed across ages: older children were more likely to acquire words with a higher exposure rate than younger children.

Around `r scales::percent(post_draws_list[["b_lv_std"]]$.rope, accuracy = 0.01)` of the posterior samples of the main effect of *Cognateness* overlapped with the ROPE completely ($\beta$ = `r round(post_draws_list[["b_lv_std"]]$.value, 3)`, 95% HDI = [`r round(post_draws_list[["b_lv_std"]]$.lower, 3)`, `r round(post_draws_list[["b_lv_std"]]$.upper, 3)`]). For every 10% increment in cognateness, the probability of word acquisition increased in `r round(post_draws_list[["b_lv_std"]]$.value/(40*sd(responses$lv)), 3)`. The effect of *Cognateness* interacted with that of *Exposure*: the 95% HDI of the regression coefficient of interaction excluded the ROPE entirely ($\beta$ = `r round(post_draws_list[["b_exposure_std:lv_std"]]$.value, 2)`, 95% HDI = [`r round(post_draws_list[["b_exposure_std:lv_std"]]$.lower, 3)`, `r round(post_draws_list[["b_exposure_std:lv_std"]]$.upper, 3)`]), suggesting that the effect of cognateness on a word's probability of acquisition changed depending on participants' exposure to the word-form. When *Exposure* was low (e.g., -1 SD), *Cognateness* increased the probability of acquisition substantially, while this effect was negligible for words with median or high exposure (+1 SD) (see @fig-marginal).

@@

This study investigated the impact of cognateness (i.e., phonological similarity between translation equivalents) on the early bilingual lexicon. We used Bayesian item response theory to model the acquisition trajectories of a large sample of Catalan and Spanish words, estimating the effect of cognateness on the probability of acquisition. This model corrected for participants' age, word-form length (number of phonemes), and a novel measure of participants' exposure rate to each word-form. Exposure rates were calculated as a language exposure-weighted lexical frequency score in which each word-form's lexical frequency was corrected by the degree to which the participant was exposed to each language. Overall, we found that cognates (i.e., phonologically similar translation equivalents) were acquired earlier than non-cognates. This effect was mediated by exposure rate: low-exposure word-forms benefited from their cognate status, whereas high-exposure word-forms did not. Using the concept of accumulator [see @kachergis2022standard for review], we provide a theoretical account of bilingual lexical acquisition. In the present account, parallel activation of the two languages plays a central role during the acquisition of early representations in the bilingual lexicon, and in which the dynamics of co-activation between translation equivalents results in an earlier age-of-acquisition.
